Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest

### 2022-07-11 article/merge.tex
first poster draft

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest},
  pdfauthor={Nicholas Medina a*; Max Villalobos b; Andrea Johnson b},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage[left]{lineno}
\linenumbers
\usepackage{setspace}\doublespace
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest}
\author{Nicholas Medina \textsuperscript{a}* \and Max Villalobos \textsuperscript{b} \and Andrea Johnson \textsuperscript{b}}
\date{scriptsize~\textsuperscript{a} Biology and Environmental Studies, Brandeis University, 415 South St, Waltham, MA, USA 02453; \textsuperscript{b} Land Stewardship, Osa Conservation, Puerto Jiménez, Golfito, Costa Rica}

@@

journals: J {[}Appl{]} Veg Sci,

@@

About half of all forests are tropical and secondary, making tropical forest regeneration integral to future forests.
Tree stand biomass and taxonomic richness can recover in a few decades, but relative abundances may lag indefinitely.
Since most forests are within a km of a habitat edge, edge effects likely affect community composition regeneration, mediated by shade tolerance and dispersal traits, along with management history.
This study addresses how wet tropical secondary forest regeneration is affected by distance to habitat edge, and hypothesizes that old timber plantations facilitate regeneration by favoring shade tolerant or late successional taxa, and adjacent forest edges mediate community composition.
An abandoned neotropical timber plantation adjacent to primary forest was censused for trees along a 300 m edge distance gradient, and analysis matched taxa to traits using relevant literature.

@@

Más de la mitad de bosques son tropicales y secundarios, indicando que la regeneración de bosques tropicales es integral al futuro de bosques, globalmente.
Biomasa arboreal y \emph{riqueza} taxonomica pueden recuperarse en algunas décadas, pero abundancias relativas pueden durar más de un siglo.
Dado que la mayoría de bosques están entre un km de su orilla, \emph{efectos de orilla} probablemente afectan la regeneración de composición de comnuidad, mediado por tolerancia a sombra y ``traits''/characterísticas de \emph{disperso}, y también historia de manejamiento.
Este estudio \emph{investiga} cómo regeneración de bosque secundario tropical húmedo es afectado por distancia a \emph{orilla de habitat}, y \emph{``hypothesizes''} que plantaciones de madera viejos facilitan regeneración porque favoran taxa tolerante a sombra o de \emph{``succession''} tarde, y que orillas de bosque cercanos/\emph{``adjacent''} \emph{median} composición de comunidad.
Una plantación de madera neotropical \emph{al lado de} bosque primario fue \emph{``censused''} por árboles siguiendo un transecto de 300 m midiendo un gradiente de distancia de orilla, y análisis combino taxa con characterísticas/\emph{``traits''} usando literatura relevante.

Mientras distancia de bosque primario aumentó, densidad de madera y \emph{``stems''} \emph{``tended to''} aumentó \emph{signifantemente}, con 10\% variación explicado, mientras tanto biomasa y luz de dócel \emph{``tended to''} quedarse igual.
\emph{Riquesa} de árboles también \emph{``tended to''} aumentar \emph{significantemente}, pero diversidad bajó profundamente y no-linearmente, y composición taxonomica varió notablemente.
Finalmente, taxa de árboles asociado \emph{etapas} de \emph{``succession''} ambos temprano y tarde bajó \emph{significativamente}, pero biomasa por modo de \emph{disperso} no \emph{``tended to''} cambiar.
\emph{Entre todo} este estudio apoya que composición es menos resiliente y más sujeto a \emph{efectos de orilla} que biomasa y \emph{riqueza}, sugeriendo que bosques globales probablemente serán \emph{asemblajes} nuevos distintos en el futuro, con \emph{``trade-offs''} entre producción de madera y biodiversidad \emph{ocuriendo} basado en actividad de manejamiento local y regional.

@@

Forest and landscape restoration is a key international conservation and climate change adaptation strategy \emph{(\protect\hyperlink{ref-depinto20}{De Pinto et al., 2020})}, and while tropical forests specifically store most land biomass \emph{(\protect\hyperlink{ref-pan13}{Pan et al., 2013})}, most are now secondary \emph{(\protect\hyperlink{ref-fao20}{FAO, 2020})} and regenerating degraded functions \emph{(\protect\hyperlink{ref-hubau20}{Hubau et al., 2020})} amid increasing climate stressors \emph{(\protect\hyperlink{ref-anderegg22}{Anderegg et al., 2022})} yet insufficient policy support \emph{(\protect\hyperlink{ref-chazdon18}{Chazdon, 2018})} via skewed priorities \emph{(\protect\hyperlink{ref-pyron22}{Pyron and Mooers, 2022})}.
Secondary forests can regenerate relatively quickly compared to old-growth forests in some ways, but in other ways remain distinct.
Neotropical secondary forests can grow quickly enough to accumulate biomass \textgreater10x faster than old-growth forests \emph{(\protect\hyperlink{ref-finegan96}{Finegan, 1996}; \protect\hyperlink{ref-poorter16}{Poorter et al., 2016})}, and substantially offset carbon emissions, with estimates ranging from \textasciitilde10\% from Amazonian deforestation \emph{(\protect\hyperlink{ref-smith20}{Smith et al., 2020})} to a decade of fossil fuel emissions from all of Latin American and the Caribbean \emph{(\protect\hyperlink{ref-chazdon16}{Chazdon et al., 2016})}.
They also recover taxonomic richness and biomass quite quickly, as well as tree height, averaging 80\% of old-growth forest levels in just 20 years, especially in wetter regions \emph{(\protect\hyperlink{ref-rozendaal19}{Rozendaal et al., 2019})}.
The recovery of species richness alone is beneficial in that it also tends to correlate with the recovery of some ecosystem services \emph{(\protect\hyperlink{ref-guariguata01}{Guariguata and Ostertag, 2001}; \protect\hyperlink{ref-stanturf14}{Stanturf et al., 2014})} like biomass and carbon storage \emph{(\protect\hyperlink{ref-liu18}{Liu et al., 2018})}.
However, secondary neotropical forests diverge widely in taxonomic composition \emph{(\protect\hyperlink{ref-norden15}{Norden et al., 2015})}, potentially taking over a century to recover \emph{(\protect\hyperlink{ref-poorter21}{Poorter et al., 2021})} with added variability \emph{(\protect\hyperlink{ref-atkinson22}{Atkinson et al., 2022})}.
Despite being one of the slowest ecosystem properties to recover, restoring recovering community composition is often important for preserving rare taxa \emph{(\protect\hyperlink{ref-carlo16}{Carlo and Morales, 2016})}, which can be keystone to locally-adapted food webs, including birds \emph{(\protect\hyperlink{ref-maas16}{Maas et al., 2016})}.
As a result, understanding secondary forest regeneration remains important\ldots{}

@@

This study was done in a neotropical lowland rainforest on the Osa Peninsula at the Greg Gund Conservation Center (approx. 8.3778176, -83.2935426) near the Piro Biological Station run by Osa Conservation.
(See \emph{Taylor et al. (\protect\hyperlink{ref-taylor15}{2015})} for a broader ecosystem description of the region.)
The study site was a regrowing 20 ha timber plantation of \emph{Bombacompsis quinata} abandoned in \textasciitilde1990 after the dry-adapted species from the Guanacaste region \emph{(\protect\hyperlink{ref-hulshof20}{Hulshof and Powers, 2020}; \protect\hyperlink{ref-perezcordero02}{Pérez Cordero and Kanninen, 2002})} grew poorly in very wet conditions.
This focal secondary forest area was roughly triangular, surrounded by primary forest on the two S and NW sides (Fig \ref{fig:map}) and a wide service road on the third NE border, with primary forest beyond it.
This census was done in 2013 during the rainy season months between June and August.
\includegraphics{merge_files/figure-latex/map-1.pdf}

@@

Edge effects were studied by dividing the secondary forest area into six 50 m strata spanning 0 - 300 m away from the primary forest edge on the S and NW sides, going inward to the E (Fig \ref{fig:map}), using available GIS software (\emph{ArcGIS 10 from 2010}, \url{esri.com}, and \emph{QGIS 2 from 2013}, \url{qgis.org}).
Each stratum was randomly filled with a number of 21 x 21 m square census plots oriented N that was proportional to its area--specifically with 11, eight, five, three, two, and one plot(s), respectively, as distance away from primary forest increased.
The total area of the 30 census plots equaled \textasciitilde1 ha or 5\% of the total secondary forest stand area, which is comparable to similar studies \emph{(\protect\hyperlink{ref-onyekwelu16}{Onyekwelu and Olabiwonnu, 2016})}.

@@

The diameter of all stems \textgreater10 cm wide were recorded in each census plot.
Following common forestry guidelines, in cases where a tree split into 2 or more stems below breast height, each stem was measured separately; in cases where a stem split only above breast height, it was measured as a single stem.
Tree height was recorded by measuring distances to both the crown and the stem at chest height \emph{(\textasciitilde2.7 m)} using a rangefinder \emph{(Bushnell, Forestry Suppliers, Inc.)} and triangulating the missing side length.
Taxa were identified with localized knowledge and field guides matching common names to Latin binomials,
and trait information was gathered from the literature.

Aboveground biomass (AGB) was estimated per stem using the allometric equation by \emph{Chave et al. (\protect\hyperlink{ref-chave15}{2015})}.
The equation specifically for wet tropical forests was used, which has performed slightly better compared to both pantropical and previous other models by \emph{(\protect\hyperlink{ref-alvarez12}{Alvarez et al., 2012}; Chave et al.; \protect\hyperlink{ref-chave05}{2005}; \protect\hyperlink{ref-ngomanda14}{Ngomanda et al., 2014})}.
The equation version with height was used since height was accessibly measurable and has shown improved local accuracy \emph{(\protect\hyperlink{ref-domke12}{Domke et al., 2012}; \protect\hyperlink{ref-khan20}{Khan et al., 2020})}.
Wood density (or specific gravity) values when available also improve biomass estimates even more than height \emph{(\protect\hyperlink{ref-vanbreugel11}{Van Breugel et al., 2011})}.
The equation used was:
\[{AGB} = 0.0673 (p D^2 H)^{0.976}\]
with diameter \emph{D} in cm, height \emph{H} in m, and wood density or specific gravity \emph{p} in \emph{g cm\textsuperscript{-3}}.
DBHs and heights were measured in the field and wood densities were extracted from available literature.
Where information was limited, values at the genus level were used and unrepresented taxa were assumed to be \emph{0.58} following World Agroforestry Database guidelines \href{worldagroforestry.org}{(worldagroforestry.org)}.

Original values found and recorded were later updated to modern values using the pipeline developed in the \emph{BIOMASS} 2.1.8 R software package by \emph{Réjou-Méchain et al. (\protect\hyperlink{ref-rejou-mechain17}{2017})}, which notably replaced unknown wood density values with plot-level averages instead during stem biomass calculations.

@@

For all response variables, data were analyzed using plot medians and subject to linear regression with distance to forest edge as the only independent variable, after assumption checks of residual normality with Shapiro-Wilk tests and equal variance with Levene tests, using base R version 4.2.0 (2022-04-22) and \emph{rstatix} 0.7.0 package functions \emph{(\protect\hyperlink{ref-base}{R Core Team, 2022})}.

Non-linear regressions were run using the function \emph{poly()} in the base R package \emph{stats} 4.2.0,
which was only ultimately recorded for stem density and richness variables, based on significance and explained variance over linear models.
All trees censused were included in each analysis including unidentified taxa that represented a single group in taxa-specific analyses.
Community analysis and ordination was done as a PERMANOVA with the \emph{adonis()} function from the \emph{vegan} 2.6.2 R package \emph{(\protect\hyperlink{ref-dixon03}{Dixon, 2003})}.
Data and code were organized with R packages \emph{here} \emph{(\protect\hyperlink{ref-here}{Müller, 2020})}, \emph{bookdown} \emph{(\protect\hyperlink{ref-bookdown2022}{Xie, 2022})}, and \emph{grateful} \emph{(\protect\hyperlink{ref-grateful}{Rodríguez-Sánchez et al., 2022})}, and are available at \url{nmedina17.github.com/osa}.

@@

Aboveground biomass did not tend to change with distance to primary forest edge
(Fig \ref{fig:biomass}a), despite significant trends with underlying related variables
(Fig \ref{fig:biomass}).

Median plot wood density tended to increase marginally significantly
(p=0.105)
with edge distance by
\ensuremath{1.9\times 10^{-4}} ±
\ensuremath{9.2\times 10^{-5}} per m
explaining
R2x100\% of variance
among median distance strata values
(Fig \ref{fig:biomass}b).

Plot stem density also tended to increase significantly
(p=0.03)
with edge distance but more strongly and non-linearly, by
\textasciitilde17.4 ±
3.4 per m
(Fig \ref{fig:biomass}c).

At this stage in forest regeneration, canopy light availability did not tend to change with distance from forest edge
(Fig \ref{fig:biomass}d).

@@

Overall diversity changed significantly with distance to primary forest edge
(Fig \ref{fig:taxa}).

Alpha diversity significantly
(p=0.003)
decreased non-linearly by
\textasciitilde0.5 ±
0 per m
as distance to primary forest edge increased
which explained
R2x100\% of variance
among distance strata medians
(Fig \ref{fig:taxa}a) --
while in contrast taxonomic richness increased slightly
(p=0.067)
and linearly
by
0 ±
0 per m
and had R2x100\% variance explained
among distance strata
(Fig \ref{fig:taxa}b).

Community composition and beta diversity also changed significantly
(p=0.01) with
11.4\%
variance among distance strata explained by distance to primary forest edge,
and the first two principal components explaining
47\% and
23\% totaling
70\%
of variance among calculated plot distances
(Fig \ref{fig:taxa}d).

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest},
  pdfauthor={Nicholas Medina},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage[left]{lineno}
\linenumbers
\usepackage{setspace}\doublespace
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Habitat edges affect tree diversity more than biomass regeneration in a wet neotropical secondary forest}
\author{Nicholas Medina}
\date{scriptsize~Ecology and Evolutionary Biology, University of Michigan, 1105 N University Ave, Ann Arbor, MI, USA 48109}

@@

@@

About half of all forests are tropical and secondary, making tropical forest regeneration integral to future forests.
Tree stand biomass and taxonomic richness can recover in a few decades, but relative abundances may lag indefinitely.
Since most forests are within a km of a habitat edge, edge effects likely affect community composition regeneration, mediated by shade tolerance and dispersal traits, along with management history.
This study addressed how wet tropical secondary forest regeneration was affected by distance to habitat edge, and hypothesized that old timber plantations facilitate regeneration by favoring shade tolerant or late successional taxa, and adjacent forest edges mediate community composition.
An abandoned neotropical timber plantation adjacent to primary forest was censused for trees along a 300 m edge distance gradient, and analysis matched taxa to traits using relevant literature.

@@

Más de la mitad de bosques son tropicales y secundarios, indicando que la regeneración de bosques tropicales es integral al futuro de bosques, globalmente.
Biomasa arboreal y riqueza taxonómica pueden recuperarse en algunas décadas, pero abundancias relativas pueden durar más de un siglo.
Dado que la mayoría de bosques quedan entre un km de su orilla, efectos de borde probablemente afectan la regeneración de composición de comunidad, mediado por tolerancia a sombra y characterísticas de disperso, y también historia de manejo.
Este estudio investigó cómo la regeneración de bosque secundario tropical húmedo fue afectado por distancia a borde de hábitat, e hipotetizó que plantaciones de madera viejas facilitan regeneración porque favorecen a taxa tolerantes a sombra o de sucesión tarde, y que orillas de bosque adyacente median cambios a composición de comunidad.
Una plantación de madera neotropical adyacente a bosque primario fue censada por árboles siguiendo un transecto de 300 m midiendo un gradiente de distancia de orilla, y análisis combinó taxa con characterísticas ecológicas usando literatura relevante.

Mientras distancia de bosque primario aumentó, densidad de madera y tallos tendieron a aumentar significativamente, con 10\% variación explicado, mientras tanto biomasa y luz de dócel tendió a quedarse igual.
Riquesa taxonómica de árboles también tendió a aumentar significativamente, pero diversidad bajó profundamente y no-linearmente, y composición taxonomica varió notablemente.
Finalmente, biomasa de taxa de árboles asociados con etapas de sucesión ambos temprano y tarde bajaron significativamente, pero biomasa por modo de disperso no tendió a cambiar.
Entre todo este estudio apoya que composición taxonómica es menos resiliente y más sujeto a efectos de borde negativos que biomasa y riqueza taxonómica, sugeriendo que bosques globales probablemente serán ensamblajes nuevamente distintos en el futuro, con compensaciones entre producción de madera y biodiversidad que ocurren basado en manejo local y regional.

@@

Forest and landscape restoration is a key international conservation and climate change adaptation strategy \emph{(\protect\hyperlink{ref-depinto20}{De Pinto et al., 2020})}, and while tropical forests specifically store most land biomass \emph{(\protect\hyperlink{ref-pan13}{Pan et al., 2013})}, most are now secondary \emph{(\protect\hyperlink{ref-fao20}{FAO, 2020})} and regenerating degraded functions \emph{(\protect\hyperlink{ref-hubau20}{Hubau et al., 2020})} amid increasing climate stressors \emph{(\protect\hyperlink{ref-anderegg22}{Anderegg et al., 2022})} yet insufficient policy support \emph{(\protect\hyperlink{ref-chazdon18}{Chazdon, 2018})} via skewed priorities \emph{(\protect\hyperlink{ref-pyron22}{Pyron and Mooers, 2022})}.
Secondary forests can regenerate relatively quickly compared to old-growth forests in some ways, but in other ways remain distinct.
Neotropical secondary forests can grow quickly enough to accumulate biomass \textgreater10x faster than old-growth forests \emph{(\protect\hyperlink{ref-finegan96}{Finegan, 1996}; \protect\hyperlink{ref-poorter16}{Poorter et al., 2016})}, and substantially offset carbon emissions, with estimates ranging from \textasciitilde10\% from Amazonian deforestation \emph{(\protect\hyperlink{ref-smith20}{Smith et al., 2020})} to a decade of fossil fuel emissions from all of Latin American and the Caribbean \emph{(\protect\hyperlink{ref-chazdon16}{Chazdon et al., 2016})}.
They also recover taxonomic richness and biomass quite quickly, as well as tree height, averaging 80\% of old-growth forest levels in just 20 years, especially in wetter regions \emph{(\protect\hyperlink{ref-rozendaal19}{Rozendaal et al., 2019})}.
The recovery of species richness alone is beneficial in that it also tends to correlate with the recovery of some ecosystem services \emph{(\protect\hyperlink{ref-guariguata01}{Guariguata and Ostertag, 2001}; \protect\hyperlink{ref-stanturf14}{Stanturf et al., 2014})} like biomass and carbon storage \emph{(\protect\hyperlink{ref-liu18}{Liu et al., 2018})}.
However, secondary neotropical forests diverge widely in taxonomic composition \emph{(\protect\hyperlink{ref-norden15}{Norden et al., 2015})}, potentially taking over a century to recover \emph{(\protect\hyperlink{ref-poorter21}{Poorter et al., 2021})} with added variability \emph{(\protect\hyperlink{ref-atkinson22}{Atkinson et al., 2022})}.
Despite being one of the slowest ecosystem properties to recover, restoring recovering community composition is often important for preserving rare taxa \emph{(\protect\hyperlink{ref-carlo16}{Carlo and Morales, 2016})}, which can be keystone to locally-adapted food webs, including birds \emph{(\protect\hyperlink{ref-maas16}{Maas et al., 2016})}.
Understanding secondary forest regeneration is key for global biodiversity conservation at the global scale, yet local management remains the strongest driver of tropical forest carbon stores \emph{(\protect\hyperlink{ref-pyles22}{Pyles et al., 2022})}.

@@

This study was done in a neotropical lowland rainforest on the Osa Peninsula at the Greg Gund Conservation Center (approx. 8.3778176, -83.2935426) near the Piro Biological Station run by Osa Conservation.
(See \emph{Taylor et al. (\protect\hyperlink{ref-taylor15}{2015})} for a broader ecosystem description of the region.)
The study site was a regrowing 20 ha timber plantation of \emph{Bombacompsis quinata} abandoned in \textasciitilde1990 after the dry-adapted species from the Guanacaste region \emph{(\protect\hyperlink{ref-hulshof20}{Hulshof and Powers, 2020}; \protect\hyperlink{ref-perezcordero02}{Pérez Cordero and Kanninen, 2002})} grew poorly in very wet conditions.
This focal secondary forest area was roughly triangular, surrounded by primary forest on the two S and NW sides (Fig \ref{fig:map}) and a wide service road on the third NE border, with primary forest beyond it.
This census was done in 2013 during the rainy season months between June and August.

\includegraphics{merge_files/figure-latex/map-1.pdf}

@@

Edge effects were studied by dividing the secondary forest area into six 50 m strata spanning 0 - 300 m away from the primary forest edge on the S and NW sides, going inward to the E (Fig \ref{fig:map}), using available GIS software (\emph{ArcGIS 10}, \url{esri.com}, and \emph{QGIS 2}, \url{qgis.org}).
Each stratum was randomly filled with a number of 21 x 21 m square census plots oriented N that was proportional to its area--specifically with 11, eight, five, three, two, and one plot(s), respectively, as distance away from primary forest increased.
The total area of the 30 census plots equaled \textasciitilde1 ha or 5\% of the total secondary forest stand area, which is comparable to similar studies \emph{(\protect\hyperlink{ref-onyekwelu16}{Onyekwelu and Olabiwonnu, 2016})}.

@@

The diameter of all stems \textgreater10 cm wide were recorded in each census plot, totaling over 1,000 trees.
Following common forestry guidelines, in cases where a tree split into 2 or more stems below breast height, each stem was measured separately; in cases where a stem split only above breast height, it was measured as a single stem.
Tree height was recorded by measuring distances to both the crown and the stem at chest height \emph{(\textasciitilde2.7 m)} using a rangefinder \emph{(Bushnell, Forestry Suppliers, Inc.)} and triangulating the missing side length.
Taxa were identified with localized knowledge and field guides matching common names to Latin binomials,
and trait information was gathered from the literature.

Aboveground biomass (AGB) was estimated per stem using the allometric equation by \emph{Chave et al. (\protect\hyperlink{ref-chave15}{2015})}.
The equation specifically for wet tropical forests was used, which has performed slightly better compared to both pantropical and previous other models by \emph{(\protect\hyperlink{ref-alvarez12}{Alvarez et al., 2012}; Chave et al.; \protect\hyperlink{ref-chave05}{2005}; \protect\hyperlink{ref-ngomanda14}{Ngomanda et al., 2014})}.
The equation version with height was used since height was accessibly measurable and has shown improved local accuracy \emph{(\protect\hyperlink{ref-domke12}{Domke et al., 2012}; \protect\hyperlink{ref-khan20}{Khan et al., 2020})}.
Wood density (or specific gravity) values when available also improve biomass estimates even more than height \emph{(\protect\hyperlink{ref-vanbreugel11}{Van Breugel et al., 2011})}.
The equation used was:
\[{AGB} = 0.0673 (p D^2 H)^{0.976}\]
with diameter \emph{D} in cm, height \emph{H} in m, and wood density or specific gravity \emph{p} in \emph{g cm\textsuperscript{-3}}.
DBHs and heights were measured in the field and wood densities were extracted from available literature.
Where information was limited, values at the genus level were used and unrepresented taxa were assumed to be \emph{0.58} following World Agroforestry Database guidelines \href{worldagroforestry.org}{(worldagroforestry.org)}.
Original values found and recorded were later updated to modern values using the pipeline developed in the \emph{BIOMASS} 2.1.8 R software package by \emph{Réjou-Méchain et al. (\protect\hyperlink{ref-rejou-mechain17}{2017})}, which notably replaced unknown wood density values with plot-level averages instead during stem biomass calculations.

Traits including successional stage and main dispersal mode were matched to species with \emph{Werden et al. (\protect\hyperlink{ref-werden20}{2020})} data, and wood tissue carbon matched to genera with the GLOWCAD database \emph{(\protect\hyperlink{ref-doraisami22}{Doraisami et al., 2022})} given high alignment at species levels \emph{(\protect\hyperlink{ref-chave06}{Chave et al., 2006})}.

@@

For all response variables, data were analyzed using plot or distance strata medians and subject to linear regression with distance to forest edge as the only independent variable, after assumption checks of residual normality with Shapiro-Wilk tests and equal variance with Levene tests, using base R version 4.2.0 (2022-04-22) and \emph{rstatix} 0.7.0 package functions \emph{(\protect\hyperlink{ref-base}{R Core Team, 2022})}.

Non-linear regressions were run using the function \emph{poly()} in the base R package \emph{stats} 4.2.0,
which was only ultimately recorded for stem density and richness variables, based on significance and explained variance over linear models.
All trees censused were included in each analysis including unidentified taxa that represented a single group in taxa-specific analyses.
Community analysis and ordination was done as a PERMANOVA with the \emph{adonis()} function from the \emph{vegan} 2.6.2 R package \emph{(\protect\hyperlink{ref-dixon03}{Dixon, 2003})}.
Data and code were organized with R packages \emph{here} \emph{(\protect\hyperlink{ref-here}{Müller, 2020})}, \emph{bookdown} \emph{(\protect\hyperlink{ref-bookdown2022}{Xie, 2022})}, and \emph{grateful} \emph{(\protect\hyperlink{ref-grateful}{Rodríguez-Sánchez et al., 2022})}, and internal pipeline \emph{oir} \emph{(\protect\hyperlink{ref-softwareMedina22}{Medina, 2022})} are stored at \url{github.com/nmedina17/osa}.

@@

Aboveground biomass did not tend to change with distance to primary forest edge
(Fig \ref{fig:biomass}a), despite significant trends with underlying related variables
(Fig \ref{fig:biomass}).
Median plot wood density tended to increase marginally significantly
(p=0.598)
with edge distance by
\ensuremath{8.2\times 10^{-5}} ±
\ensuremath{1.4\times 10^{-4}} per m
explaining
R2x100\% of variance
among median distance strata values
(Fig \ref{fig:biomass}b).
Plot stem density also tended to increase significantly
(p=0.82)
with edge distance but more strongly and non-linearly, by
\textasciitilde-65.9 ±
78.9 per m
(Fig \ref{fig:biomass}c).
At this stage in forest regeneration, canopy light availability did not tend to change with distance from forest edge
(Fig \ref{fig:biomass}d).

@@

Overall diversity changed significantly with distance to primary forest edge
(Fig \ref{fig:taxa}).
Diversity significantly
(p=0.414)
decreased non-linearly by
\textasciitilde0.1 ±
0.1 per m
as distance to primary forest edge increased,
which explained
R2x100\% of variance
among distance strata medians
(Fig \ref{fig:taxa}a) --
while in contrast taxonomic richness increased slightly
(p=0.067)
and linearly
by
0 ±
0 per m
and had R2x100\% variance explained
among distance strata
(Fig \ref{fig:taxa}b).
Community composition and beta diversity also changed significantly
(p=0.01) with
25.6\%
variance among distance strata explained by distance to primary forest edge,
and the first two principal components explaining
59\% and
24\% totaling
83\%
of variance among calculated plot distances
(Fig \ref{fig:taxa}d).
Key abundant taxa \emph{Vochysia} and \emph{Ficus} showed different responses -- \emph{Vochysia} nearly tended to decrease with distance to primary forest edge
(Fig \ref{fig:taxa}e),
while \emph{Ficus} decreased marginally significantly
(p=0.088)
and linearly by
-62.6 ±
25.1 per m
with R2x100\% variance among distance strata explained
(Fig \ref{fig:taxa}f).
